Modal type theory based on the intuitionistic epistemic logic

### 2018-01-28 article.tex
added proof
added confluence proof

  6) ${\bf let \: pure \:} \underline{\quad} = \underline{\quad} {\: \bf in \:} N \rightarrow_{\beta} {\bf pure \:} N$;

  7) $\lambda x. f x \rightarrow_{\eta} f$;

  8) $\langle \pi_1 P, \pi_2 P \rangle \rightarrow_{\eta} P$;

  9) ${\bf let \: pure \:} x = M {\: \bf in \: } x \rightarrow_{\eta} M$;

  10) $M \rightarrow_{\beta \eta} N \Rightarrow {\bf pure \: M} \rightarrow_{\beta \eta} {\bf pure \: N}$
  \end{defin}

@@

\begin{lemma} Basic lemmas for ${\bf K}_I$.

  i) Let $\Gamma \vdash {\bf pure \:} M : {\bf K}A$, then $\Gamma \vdash M : A$;

  ii) Let $\Gamma \vdash M : A$ and $\Gamma \subseteq \Delta$, then $\Delta \vdash M : A$;

  iii) Let $\Gamma, x : A \vdash {\bf pure \:} M : {\bf K}B$ and $\Gamma \vdash N : A$, then $\Gamma \vdash {\bf pure \:} M [x := N] : {\bf K}B$.

  iv) Let $\Gamma \vdash {\bf pure \:} M : {\bf K}A$, then $\Gamma [B := C] \vdash {\bf pure \:} M : {\bf K} (A [B := C])$.
\end{lemma}

@@

  i) Induction on $\Gamma \vdash {\bf pure \:} M : {\bf K}A$ and $\Gamma \vdash {\bf let \: pure \:} \vec{x} = \vec{N} {\: \bf in \:} N : {\bf K}B$ correspondently.

  ii) Let $\Gamma \vdash {\bf pure \:} M : {\bf K}A$. Then $\Gamma \vdash M : A$ by generation and $\Delta \vdash M : A$ by assumption. So $\Delta \vdash {\bf pure \:} M : {\bf K} A$ by ${\bf K}_I$.

  iii) Let $\Gamma, x : A \vdash {\bf pure \:} M : {\bf K}B$ and $\Gamma \vdash N : A$.

  By generation $\Gamma, x : A \vdash M : B$ and by assumption $\Gamma \vdash M [x := N] : B$.

  By $K_I$, $\Gamma \vdash {\bf pure \:} (M [x := N]) : {\bf K}B$.

  iv) Let $\Gamma \vdash {\bf pure \:} M : {\bf K}A$. By generation $\Gamma \vdash M : A$ and by assumption $\Gamma [B := C] \vdash M : A[B := C]$.

  By $K_I$ $\Gamma \vdash {\bf pure \:} M : {\bf K}(A[B := C])$.
\end{proof}

@@

  i) $\twoheadrightarrow_{\beta}$ is strongly normalizing;

  ii) $\twoheadrightarrow_{\beta}$ is confluent.
\end{theorem}

@@

It is easy to extend Kakutani's result for ${\bf pure}$-rules.

ii) Confluence can be proved by extending Barendregt's technique with term underlying for untyped or simply typed lambda calculus \cite{Baren}.

Hence, it is sufficient to prove strip lemma.
\end{proof}

@@

  Normal form in call-by-name $\lambda_{{\bf K}}$ has the subformula property.

@@

2) $[\![\vdash {\bf let \: pure \:} \underline{\quad} = \underline{\quad} {\: \bf in \:} N : {\bf }K A ]\!] = [\![\vdash {\bf pure \:} N : {\bf K A}]\!]$.

$\begin{array}{lll}
&[\![\vdash {\bf let \: pure \:} \underline{\quad} = \underline{\quad} {\: \bf in \:} N : {\bf }K A ]\!] = & \text{By interpetation}\\
&\mathcal{K}([\![N]\!]) \circ \eta_{\mathds{1}} =& \text{Naturality for $\eta$} \\
&\eta_{[\![A]\!]} \circ [\![N]\!] =& \text{By interpretation} \\
&[\![\vdash {\bf pure \:} N : {\bf K} A]\!]&
\end{array}$

3)

If $\Gamma \vdash M : A$ and $M \rightarrow_{\beta \eta} N$, then $[\![\Gamma \vdash {\bf pure \:} M : {\bf K}A]\!] = [\![\Gamma \vdash {\bf pure \:} N : {\bf K}A]\!]$.

If $\Gamma \vdash M : A$ and $M \rightarrow_{\beta \eta} N$, then $\Gamma \vdash N : A$ by subject reduction.

By assumption $[\![\Gamma \vdash M : A]\!] = [\![\Gamma \vdash N : A]\!]$.

So $\eta_{[\![A]\!]} \circ [\![\Gamma \vdash M : A]\!] = \eta_{[\![A]\!]} \circ [\![\Gamma \vdash N : A]\!]$.

Hence $[\![\Gamma \vdash {\bf pure \:} M : {\bf K}A]\!] = [\![\Gamma \vdash {\bf pure \:} N : {\bf K}A]\!]$.

@@

Follows from i) and ii).

@@

@@

  %далее сам список используевой литературы
  \begin{thebibliography}{}

>>>>

  6) $\lambda x. f x \rightarrow_{\eta} f$;

  7) $\langle \pi_1 P, \pi_2 P \rangle \rightarrow_{\eta} P$;

  8) ${\bf let \: pure \:} x = M {\: \bf in \: } x \rightarrow_{\eta} M$;
  \end{defin}

By default we use call-by-name evaluation strategy.

@@

\begin{lemma} Generation for ${\bf K}_I$.

  Let $\Gamma \vdash {\bf pure \:} M : {\bf K}A$, then $\Gamma \vdash M : A$;
\end{lemma}

\begin{proof}
  Induction on the structure of ${\bf pure \:} M$.
\end{proof}

\begin{lemma} Basic lemmas .

  i) Let $\Gamma \vdash M : A$ and $\Gamma \subseteq \Delta$, then $\Delta \vdash M : A$;

  ii) Let $\Gamma, x : A \vdash M : B$ and $\Gamma \vdash N : A$, then $\Gamma \vdash M [x := N] : B$.

  iii) Let $\Gamma \vdash M : A$, then $\Gamma [B := C] \vdash M : (A [B := C])$.
\end{lemma}

@@

  i-ii-iii) Induction on $\Gamma \vdash M : A$.

@@

  $\twoheadrightarrow_{\beta}$ is strongly normalizing;
\end{theorem}

@@

@@

  $\twoheadrightarrow_{\beta}$ is confluent.
\end{theorem}

  We modify and apply Barendregt's technique with term underlying. We will consider the fragment of the grammar for terms without
  constructors for pairs for simplicity.

\begin{defin} The set of underlined terms.

  \begin{itemize}
    \item $x \in \mathbb{V} \Rightarrow x \in \underline{\Lambda}$;
    \item $M \in \underline{\Lambda} \Rightarrow (\lambda x. M) \in \underline{\Lambda}$;
    \item $M, N \in \underline{\Lambda} \Rightarrow (M N) \in \underline{\Lambda}$;
    \item $M \in \underline{\Lambda} \Rightarrow ({\bf pure \:} M) \in \underline{\Lambda}$;
    \item $\vec{x} \in \mathbb{V}, \vec{M}, N \in \underline{\Lambda} \Rightarrow {\bf let \: pure \:} \vec{x} = \vec{M} {\: \bf in \:} N \in \underline{\Lambda}$;
    \item $M, N \in \underline{\Lambda} \Rightarrow (\lambda_i x. M) N \in \underline{\Lambda}$, for all $i \in \mathbb{N}$.
  \end{itemize}
\end{defin}

\begin{defin} Subsitution for term with labelled lambda:

  $((\lambda_i x. M) N) [y := Z] = (\lambda_i x. M [y := Z]) (N [y := Z])$
\end{defin}

\begin{defin} Index erasing

  Let us define map $|.| : \underline{\Lambda} \to \Lambda$ as follows:

  \begin{itemize}
    \item $|x| = x$;
    \item $|\lambda x. M| = \lambda x. |M|$;
    \item $|M N| = |M| |N|$;
    \item $|{\bf pure \:} M| = {\bf pure \:} |M|$;
    \item $|{\bf let \: pure \:} \vec{x} = \vec{M} {\: \bf in \:} N| = {\bf let \: pure \:} \vec{x} = \vec{|M|} {\: \bf in \:} |N|$;
    \item $|(\lambda_i x. M) N| = (\lambda x. M) N$
  \end{itemize}
\end{defin}

\begin{defin} Reduction rules:

\begin{itemize}
  \item $(\lambda x. M) N \rightarrow_{\underline{\beta}} M [x := N]$;
  \item $\begin{array}{llll}
  & {\bf let \: pure \:} \langle \vec{x}, y, \vec{z} \rangle = \langle \vec{M}, {\bf let \: pure \: } \vec{w} = \vec{N} {\bf \: in \: } Q, \vec{P} \rangle {\: in \:} R \rightarrow_{\underline{\beta}} & \\
  & {\bf let \: pure \:} \langle \vec{x}, \vec{w}, \vec{z} \rangle = \langle \vec{M}, \vec{N}, \vec{P} \rangle \: { \bf in } \: R [y := Q]
  \end{array}$;
  \item ${\bf let \: pure \:} \vec{x} = {\bf pure \:} \vec{M} {\: \bf in \:} N \rightarrow_{\underline{\beta}} {\bf pure \:} N [\vec{x} := \vec{M}]$;
  \item $(\lambda x_i. M) N \rightarrow_{\underline{\beta}} M [x := N]$
\end{itemize}
\end{defin}

$\twoheadrightarrow_{\underline{\beta}}$ is a reflexive-transitive closure of $\rightarrow_{\underline{\beta}}$.

\begin{defin} Indexed redex erasing:

  Let us define the next map $\phi : \underline{\Lambda} \to \Lambda$:

  \begin{itemize}
    \item $\phi(x) = x$;
    \item $\phi(\lambda x. M) = \lambda x. \phi(M)$;
    \item $\phi(M N) = \phi(M) \phi(N)$;
    \item $\phi({\bf pure \:} M) = {\bf pure \:} \phi(M)$;
    \item $\phi({\bf let \: pure \:} \vec{x} = \vec{M} {\: \bf in \:} N) = {\bf let \: pure \:} \vec{x} = \vec{\phi(M)} {\: \bf in \:} \phi(N)$;
    \item $\phi((\lambda_i x. M) N) = M [x := N]$
  \end{itemize}
\end{defin}

  $\forall \underline{M}, \underline{N} \in \underline{\Lambda} \: \forall M, N \in \Lambda, \text{if } |\underline{M}| = M, |\underline{N}| = N, \text{then}$
  \begin{itemize}
    \item If $M \twoheadrightarrow_{\beta} N$, then $\underline{M} \twoheadrightarrow_{\underline{\beta}} \underline{N}$
    \item Vice versa
  \end{itemize}
\end{lemma}

  Induction on the generation $\rightarrow_{\beta}$ and $\rightarrow_{\underline{\beta}}$ correspondently. The general statement follows from transitivity
  of multi-step reductions of both types.
\end{proof}

\begin{lemma}
  $\phi(M [x := N]) = \phi(M) [x := \phi(N)]$.
\end{lemma}

\begin{proof}
  We treat only cases with {\bf pure} and with {\bf let}. For the rest cases see \cite{Baren}.

  1)
  $ $

  $\begin{array}{lll}
  & \phi({\bf pure \:} (M [x := N])) = & \text{By the definition of $\phi$}\\
  & {\bf pure \:} (\phi(M [x := N])) = & \text{Induction hypothesis} \\
  & {\bf pure \:} (\phi(M) [x := \phi(N)]) = & \text{Substitution definition}\\
  & ({\bf pure \:} \phi (M)) [x := \phi(N)]&
  \end{array}$

  2)
  $ $

  $\begin{array}{lll}
  &\phi(({\bf let \: pure \:} \vec{x} = \vec{M} {\: \bf in \:} N) [y := P]) = & \text{Substitution definition}\\
  &\phi({\bf let \: pure \:} \vec{x} = (\vec{M}[y := P]) {\: \bf in \:} N) = & \text{By the definition of $\phi$}\\
  &{\bf let \: pure \:} \vec{x} = \phi(\vec{M}[y := P]) {\: \bf in \:} \phi(N) = & \text{Induction hypothesis} \\
  &{\bf let \: pure \:} \vec{x} = (\phi(\vec{M}) [y := \phi(P)]) {\: \bf in \:} \phi(N) =& \text{Substitution definition} \\
  &({\bf let \: pure \:} \vec{x} = \phi(\vec{M}) {\: \bf in \:} \phi(N))[y := \phi(P)]
  \end{array}$

  \begin{itemize}
  \item If $M \twoheadrightarrow_{\underline{\beta}} N$, then $\phi(M) \twoheadrightarrow_{\beta} \phi(N)$
  \item If $|M| = N$ and $\phi(M) = P$, then $N \twoheadrightarrow_{\beta} P$.
  \end{itemize}
\end{lemma}

  i) Induction on the generation of $\twoheadrightarrow_{\underline{\beta}}$ using previous lemma.

  ii) Induction on the structure of $M$.
\end{proof}

\begin{lemma} Strip lemma.

If $M \rightarrow_{\beta} N$ and $M \twoheadrightarrow_{\beta} P$. Then there exists some term $Q$, such that
$N \twoheadrightarrow_{\beta} Q$ and $P \twoheadrightarrow_{\beta} Q$.
\end{lemma}

Proof is similar to \cite{Baren} \cite{Baren2}. We build the following diagram, which commutes by lemmas 5 and 7.

\xymatrix{
&&& M \ar[dd]_{\beta} \ar@{->>}[rrr] &&& P \ar@{-->>}[dd]^{\beta} \\
&&&& N' \ar[ul]^{|.|} \ar[dl]_{\phi} \ar@{-->>}[rrr]^{\underline{\beta}} &&& P' \ar[ul]^{|.|} \ar[dl]_{\phi} \\
&&& N \ar@{-->>}[rrr]_{\beta} &&& Q
}
\end{proof}

\begin{col}
$ $
If $M \twoheadrightarrow_{\beta} N$ and $M \twoheadrightarrow_{\beta} P$. Then there exists some term $Q$, such that
$N \twoheadrightarrow_{\beta} Q$ and $P \twoheadrightarrow_{\beta} Q$.
\end{col}

  Unfold $M \twoheadrightarrow_{\beta} N$ as the sequence of one-step reduction and apply strip lemma on the every step.
\end{proof}

  Normal form in $\lambda_{{\bf K}}$ has the subformula property.

@@

@@

$\begin{array}{lll}
& \ast_{A,B} \circ (\eta_{A} \times \eta_{B}) = & \\
& \text{By unfolding} & \\
& [q, {\bf let \: pure \:} x, y = \pi_1 q, \pi_2 q {\: \bf in \:} \langle x, y \rangle] \circ [p, \langle {\bf pure \:} (\pi_1 p), {\bf pure \:} (\pi_2 p) \rangle] = & \\
& \text{Composition}& \\
& [p, {\bf let \: pure \:} x, y = \pi_1 q, \pi_2 q {\: \bf in \:} \langle x, y \rangle [q := \langle {\bf pure \:} (\pi_1 p), {\bf pure \:} (\pi_2 p) \rangle]] = & \\
& \text{By substitution}& \\
& [p, {\bf let \: pure \:} x, y = \pi_1 (\langle {\bf pure \:} (\pi_1 p), {\bf pure \:} (\pi_2 p) \rangle), \pi_2 (\langle {\bf pure \:} (\pi_1 p), {\bf pure \:} (\pi_2 p) \rangle) {\: \bf in \:} \langle x, y \rangle] = & \\
& \text{Reduction rules}& \\
& [p, {\bf let \: pure \:} x, y = {\bf pure \:} (\pi_1 p), {\bf pure \:} (\pi_2 p) {\: \bf in \:} \langle x, y \rangle] = & \\
& \text{Reduction rule}& \\
& [p, {\bf pure \:} (\langle x,y \rangle [x := \pi_1 p, y := \pi_2 p])] = & \\
& \text{Substitution}& \\
& [p, {\bf pure \:} \langle \pi_1 p, \pi_2 p \rangle] = & \\
& \text{$\eta$-reduction}& \\
& [p, {\bf pure \:} p] =& \\
& \text{By definition}& \\
& \eta_{A \times B}&
\end{array}$

@@

\section{Monadic extention}

In modern Haskell \verb"Monad" class is an extention of the \verb"Applicative" class,
where monadic binding (\verb"(>>=)") is the single basic class method:

\begin{lstlisting}[language=Haskell]
  class Applicative m => Monad (m :: * -> *) where
    (>>=) :: m a -> (a -> m b) -> m b
    (>>) :: m a -> m b -> m b
    return :: a -> m a
    fail :: String -> m a
\end{lstlisting}
From a logical point of view we may consider this extention as the extention of IEL$^{-}$ with the imdepotency axiom.

\begin{defin} ``Monadic'' intuitionistic epistemic logic IEL$^{-}$

  4) ${\bf KK} A \to {\bf K}A$

  Rule: MP.
\end{defin}

Categorically \verb"Monad" class is a Kleisli triple on monoidal category, or, which is the same, strong monoidal monad:

\begin{defin} Strong monoidal monad

\begin{defin} Monadic $\lambda_{{\bf K}}$

  Monadic $\lambda_{{\bf K}}$ is a $\lambda_{{\bf K}}$ with the next additional typing rule:

  \begin{prooftree}
    \AxiomC{$\Gamma \vdash M : {\bf KK}A$}
    \RightLabel{${\bf K}^2$--e}
    \UnaryInfC{$\Gamma \vdash {\bf join \:} M : {\bf K}A$}
  \end{prooftree}
\end{defin}

\begin{defin}
\end{defin}

@@

  \begin{thebibliography}{}
